Dynamic Modeling, Parameter Estimation and Uncertainty Analysis in \proglang{R}

### 2016-09-26 article.tex
Anfang der Einleitung

\noindent In this paper, we present the \pkg{dMod} package which provides an \proglang{R} framework to deal with the inverse problem in dynamic systems. The particularity of the \pkg{dMod} approach is to provide and propagate accurate derivatives computed from symbolic expressions. This derivative information highly supports the convergence of optimization routines and provides the numerical stability required by sofisticated uncertainty analysis methods. Computational efficiency is guaranteed by automatic generation and integration of \proglang{C} code. The framework is object oriented (S3) and provides a variety of functions to set up dynamic models, observation functions and parameter transformations for multi-conditional parameter estimation.

\noindent The key elements of the framework and the methodology implemented in \pkg{dMod} are highlighted by an application on a three-compartment transporter model.

@@

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
\section{Introduction}
	Dynamic models are found in various research fields, ranging from physics to biology or finance. In a dynamic model, the single terms correspond to mechanisms or processes which are combined to a whole, giving rise to a complex system. If the single terms are associated to parameters, parameter estimation can identify those processes of the complex system that are crucial to explain the observation. In that sense, parameter estimation can

@@

There are several ways to define the differential equations. An \code{eqnvec} of equations can be explicitly formulated, analogously to a \code{c()} command. Especially for chemical reactions it can be tideous to keep track of all gain and loss terms. Therefore, \pkg{dMod} provides the \code{eqnlist} class which encodes the ODE by a list of (1) state names, (2) rate expressions, (3) compartment volumes and (4) the stoichiometric matrix. Each reaction flux occurs only once and the gain and loss is represented by the coefficients in the stoichiometric matrix [REF]. \pkg{dMod} supports the read-in of a csv file with the stoichiometric matrix and rate expressions and provides a function \code{addReaction()} to construct an \code{eqnlist} object step-by-step and add further reactions to an already existing \code{eqnlist}. The conversion to the (differential) equations is done by \code{as.eqnvec()} which does all the bookkeeping of gain and loss terms and volume ratios due to compartment transitions.

@@

The \pkg{dMod} package allows to define transformations $\Phi$, see eq.~\eqref{eq:partrafo}, by the \code{"+"} operator: 
\begin{equation*}
\verb!p <- P(eqnvec1, condition = "one") + P(eqnvec2, condition = "two")!. 
\end{equation*}
All symbols from \code{eqnvec1} and \code{eqnvec2} are collected and their union constitutes the symbols of $\theta$. The evaluation \code{p(theta)} returns a list of length two with inner parameters. The \code{"+"} operator can be applied consecutively to add condition by condition.

@@

The aim of \pkg{dMod} is parameter estimation. The objective function to be minimized for this purpose is the weighted least squares function which is produced by the command \code{normL2(data, prdfn)} where \code{data} is a \code{datalist} object and \code{prdfn} is a prediction function. A \code{datalist} is a list of \code{data.frame}s providing time points, observables, measured values and uncertainties for each condition. The objective function is the final link in the chain of parameter transformations, prediction- and observation functions. It collects all derivative information and besides the objective value it computes gradient and Hessian. The standard optimizer employed within \pkg{dMod} is the \code{trust()} optimizer from the \pkg{trust} package.

Objective functions can be added by the \code{"+"} operator, meaning the objective values, gradients and Hessians are added. Besides \code{normL2()}, functions like \code{constraintL2()} and \code{datapointL2()} produce objective functions, too. They allow to define parameter priors or treat data points as parameters.

@@

>>>>

In this paper, we present the \proglang{R} package \pkg{dMod} which provides a  framework to deal with the inverse problem in dynamic systems. The particularity of the \pkg{dMod} approach is to provide and propagate accurate derivatives computed from symbolic expressions wherever possible. This derivative information highly supports the convergence of optimization routines and provides the numerical stability required by sofisticated uncertainty analysis methods. Computational efficiency is guaranteed by automatic generation and integration of \proglang{C} code. The framework is object oriented (S3) and provides a variety of functions to set up dynamic models, observation functions and parameter transformations for multi-conditional parameter estimation.

The key elements of the framework and the methodology implemented in \pkg{dMod} are highlighted by an application on a three-compartment transporter model within this paper.

@@

%% include your article here, just as usual
%% Note that you should use the \pkg{}, \proglang{} and \code{} commands.
\section{Introduction}
	Dynamic models are found in several research fields, ranging from physics to biology or finance. In all these fields, models are used to construct a link between theoretical concepts and empirical evidence. In a dynamic model, the single terms correspond to mechanisms or processes which are combined to a whole, giving rise to a complex system. If the single terms are associated to parameters, parameter estimation can identify those processes of the complex system that are crucial to explain the observation. In that sense, parameter estimation can be employed as an instrument to understand complex systems. Once, the link between observation and model is established, questions about the identifiability of parameters arise. The parameter space needs to be explored to see whether the estimate is unique and to determine confidence bounds.

	The key requirement for parameter estimation is accurate and robust non-linear optimization. Many of the standard optimization routines implemented in \proglang{R} use derivative information. Indeed, if the objective function is sufficiently smooth, derivatives can considerably reduce the number of iterations and convergence is more reliable [REF]. However, the computation of derivatives in the context of ODE models holds some pit-falls [REF] which are e.g.~related to the step-size control of the integrator.

@@

There are several ways to define the differential equations. An \code{eqnvec} of equations can be explicitly formulated, analogously to a \code{c()} command. Especially for chemical reactions it can be tideous to keep track of all gain and loss terms. Therefore, \pkg{dMod} provides the \code{eqnlist} class which encodes the ODE by a list of (1) state names, (2) rate expressions, (3) compartment volumes and (4) the stoichiometric matrix. Each reaction flux occurs only once and the gain and loss is represented by the coefficients in the stoichiometric matrix [REF]. \pkg{dMod} supports the read-in of a csv file with the stoichiometric matrix and rate expressions and provides a function \code{addReaction()} to construct an \code{eqnlist} object step-by-step or add further reactions to an already existing \code{eqnlist}. The conversion to the (differential) equations is done by \code{as.eqnvec()} which does all the bookkeeping of gain and loss terms and volume ratios due to compartment transitions.

@@

The \pkg{dMod} package allows to define transformations $\Phi$, see eq.~\eqref{eq:partrafo}, by the \code{"+"} operator: 
\begin{equation*}
\verb!p <- P(eqnvec1, condition = "one") + P(eqnvec2, condition = "two")!. 
\end{equation*}
All symbols from \code{eqnvec1} and \code{eqnvec2} are collected and their union constitutes the symbols of $\theta$. The evaluation \code{p(theta)} returns a list (of length two in the example) with inner parameters. The \code{"+"} operator can be applied consecutively to add condition by condition.

@@

The aim of \pkg{dMod} is parameter estimation. The objective function to be minimized for this purpose is the weighted least squares function which is produced by the command \code{normL2(data, prdfn)} where \code{data} is a \code{datalist} object and \code{prdfn} is a prediction function. A \code{datalist} is a list of \code{data.frame}s providing time points, observables, measured values and uncertainties for each condition. The \code{as.datalist()} command allows to construct a datalist from a \code{data.frame} as it is typically used for mixed effects models or plotting with \pkg{ggplot2}.

The objective function is the final link in the chain of parameter transformations, prediction- and observation functions. It collects all derivative information and besides the objective value it computes gradient and Hessian. The standard optimizer employed within \pkg{dMod} is the \code{trust()} optimizer from the \pkg{trust} package.

Objective functions can be added by the \code{"+"} operator, meaning the objective values, gradients and Hessians are accumulated by summation. Besides \code{normL2()}, functions like \code{constraintL2()} and \code{datapointL2()} produce objective functions, too. They allow to define parameter priors or treat data points as parameters.

@@

\subsection{Prediction uncertainty and validation profiles}
